---
title: "Assignment - # 5"
author: Mojisola Popoola
date: "`r Sys.Date()`"
format: html
editor: visual
theme: cosmo
---

This is a Behavioral Risk Factor data on Tobacco use from 2011-2019. A Centers for Disease Control and Prevention (CDC), State Tobacco Activities Tracking and Evaluation (STATE) System. The Behavioral Risk Factor Surveillance System (BRFSS) is a continuous, state-based surveillance system that collects information about modifiable risk factors for chronic diseases and other leading causes of death. The data for the STATE System were extracted from the annual BRFSS surveys from participating states. Tobacco topics included are cigarette and e-cigarette use prevalence by demographics, cigarette and e-cigarette use frequency, and quit attempts. Data link [CDC Data](https://dev.socrata.com/foundry/data.cdc.gov/wsas-xwh5).

# load libraries

```{r, message=FALSE, warning=FALSE}
library(haven)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(summarytools)
```

```{r}
rm(list = ls())
```

# loading data

```{r}
Tobacco_Use <- read.csv("Tobacco_Use.csv")
```

# Exploratory/Descriptive Analysis

```{r}
my_data <- Tobacco_Use %>% 
  select(Year = YEAR, Abbr = LocationAbbr, State = LocationDesc, MeasureDesc, `Data_Value %` = Data_Value, Gender, Race, Age, Education)
```

# Remove any missing values if needed

```{r}
my_data<-na.omit(my_data)
anyNA(my_data)
```

```{r}
str(my_data) 
```

# Cleaning the data

# Change character variables to factor

```{r}
my_data$Year <- as.factor(my_data$Year)

str(my_data) 
```

# Summary Statistics

```{r}
summary(my_data)

```

# Distribution of Numeric Variables

```{r}
# Histogram for 'Data_Value %'
ggplot(my_data, aes(x = `Data_Value %`)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Data Value in %", x = "Data Value", y = "Frequency")
```

```{r}
# Convert Age to a factor for categorical analysis
my_data$Age <- as.factor(my_data$Age)
summary(my_data$Age)

# Load necessary library for visualization
library(ggplot2)

# Bar plot of Age distribution
ggplot(my_data, aes(x = Age)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  coord_flip() 
  
```

# Relationships Between Variables

```{r}
# Scatter plot between Year and Data_Value %
ggplot(my_data, aes(x = Year, y = `Data_Value %`)) +
  geom_point() +
  labs(title = "Scatter Plot of Year vs Data Value in %", x = "Year", y = "Data Value %") +
   coord_flip() 

# Box plot for Data_Value % by Gender
ggplot(my_data, aes(x = Gender, y = `Data_Value %`)) +
  geom_boxplot() +
  labs(title = "Box Plot of Data Value % by Gender", x = "Gender", y = "Data Value %")

# Box plot for Data_Value % by Race
ggplot(my_data, aes(x = Race, y = `Data_Value %`)) +
  geom_boxplot() +
  labs(title = "Box Plot of Data Value % by Race", x = "Race", y = "Data Value %") +
   coord_flip() 
```

# Table 1. Descriptive Statistics

```{r}
library(dplyr)
library(gtsummary)

# Select relevant columns
my_data %>%
  dplyr::select(all_of(c("Year", "Age","Gender","Race", "MeasureDesc", "Education"))) -> my_data_baseline

# Create summary table
my_data_baseline %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"),
    missing_text = "(Missing)"
  )
```

# plot to seeif it's approximately normal,

```{r}
# Histogram of the continuous variable
ggplot(my_data, aes(x = `Data_Value %`)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Histogram of Data Values", x = "Data_Value %", y = "Frequency")

# Q-Q plot of the continuous variable
qqnorm(my_data$`Data_Value %`, main = "Q-Q Plot of Data Values")
qqline(my_data$`Data_Value %`, col = "red")

# Density plot of the continuous variable
ggplot(my_data, aes(x = `Data_Value %`)) +
  geom_density(fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Density Plot of Data Values", x = "Data_Value %", y = "Density")
```

The graphs show tha the distributions are not normally distributed due to the fact that the Q-Q plot has few points at the ends that deviate significantly from the line, indicating potential outliers or deviations in the tails.

## Amelia Tristan contributed to this exercise.

```{r}
set.seed(1000)
n = 1200

data_val = round(runif(n, 0, 100), 2)

#creating smoking data set with age, year, gender, race, education, and smoking status
age_levels = c("18 to 24 Years", "All Ages", "Age 25 and Older", "Age 20 and Older", "65 Years and Older", "45 to 64 Years", "25 to 44 Years", "18 to 44 Years")
year_levels = c("2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019")
gender_levels = c("Male", "Female", "Overall")
race_levels = c("All Races", "White", "Hispanic", "Asian/Pacific Islander", "African American", "American Indian/Alaska Native")
education_levels = c("All Grades", "> 12th Grade", "< 12th Grade", "12th Grade")
smoke_status = c("Current", "Never", "In the past year", "In the past 3 months", "In the past month", "Socially", "Occasionally")

#creating sample
age = sample(age_levels, n, replace = TRUE, prob = c(round(runif(8, 0, 1),2)))
year = sample(year_levels, n, replace = TRUE, prob = c(round(runif(9, 0, 1),2)))
gender = sample(gender_levels, n, replace = TRUE, prob = c(round(runif(3, 0, 1),2)))
race = sample(race_levels, n, replace = TRUE, prob = c(round(runif(6, 0, 1),2)))
education = sample(education_levels, n, replace = TRUE, prob = c(round(runif(4, 0, 1),2)))
smoking = sample(smoke_status, n, replace = TRUE, prob = c(round(runif(7, 0, 1),2)))

tob_fake = data.frame(age, year, gender, race, education, smoking, data_val)
summary(tob_fake)
```
```{r}
# Histogram for data_val
ggplot(tob_fake, aes(x = data_val)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Data Value in %", x = "Data Value", y = "Frequency")
```
The data values are more evenly distributed than the original data with no apparent skewed distribution(normal).

```{r}
#age to factor as Moji did
tob_fake$age <- as.factor(tob_fake$age)

# Bar plot of Age distribution
ggplot(tob_fake, aes(x = age)) +
  geom_bar() +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  coord_flip() 

```
The same can be said for age (normally distributed). All ages has the third highest frequency, unlike the original data (all ages was the top age range).

```{r}
# Box plot for data_val by Gender
ggplot(tob_fake, aes(x = gender, y = data_val)) +
  geom_boxplot() +
  labs(title = "Box Plot of Data Value by Gender", x = "Gender", y = "Data Value %")

# Box plot for data_val by Race
ggplot(tob_fake, aes(x = race, y = data_val)) +
  geom_boxplot() +
  labs(title = "Box Plot of Data Value by Race", x = "Race", y = "Data Value %")

#bar plot of smoking status by race
ggplot(tob_fake, aes(x = race)) + geom_bar() + facet_wrap(~smoking) + labs(x = "Race", y = "Smoking Status", title = "Smoking Status Frequency by Race") + coord_flip()

#bar plot of smoking status by education
ggplot(tob_fake, aes(x = education)) + geom_bar() + facet_wrap(~smoking) + labs(x = "Education Level", y = "Smoking Status", title = "Smoking Status Frequency by Education Level") + coord_flip()
```
Data values have little variance by gender and race. Race is normally distributed by smoking status across all levels. Education level is skewed by smoking status across all levels.
```{r}
# Q-Q plot of the continuous variable
qqnorm(tob_fake$data_val, main = "Q-Q Plot of Data Values")
qqline(tob_fake$data_val, col = "red")
```
The plot is not normally distributed and is actually similar to the original data's qqplot. This contradicts the histogram of data values that suggests it is normally distributed.

