---
title: "cdcdata-exercise"
format: pdf
editor: visual
---

## ##2015-2017. High School Dataset – Including Sexual Orientation. The Youth Risk Behavior Surveillance System (YRBSS) monitors six categories of priority health behaviors among youth and young adults: 

## 1) behaviors that contribute to unintentional injuries and violence; 2) tobacco use; 

## 3) alcohol and other drug use; 

## 4) sexual behaviors related to unintended pregnancy and sexually transmitted infections (STIs), including human immunodeficiency virus (HIV) infection; 

## 5) unhealthy dietary behaviors; and 

## 6) physical inactivity. In addition, YRBSS monitors the prevalence of obesity and asthma and other priority health behaviors. This dataset contains national, state, and local data from 2015 that includes two aspects of sexual orientation – sexual identity and sex of sexual contacts. Additional information about the YRBSS can be found at [www.cdc.gov/yrbss](http://www.cdc.gov/yrbss).

<https://data.cdc.gov/Youth-Risk-Behaviors/DASH-YRBSS-HS-Cigar-Use-by-total-grade-sex-race-se/bedg-mmpy/about_data>

#load the data into R

```{r}
library(dplyr)
library(tidyverse)
library(caret)
```

```{r}
youth=read_csv("/Users/knowledgedivine/Desktop/practicum2/mojisolapopola-P2-portfolio/cdcdata-exercise/youth.csv")
```

```{r}
str(youth)

```

Remove missing value

```{r}
df_youth <- youth[, colSums(is.na(youth)) == 0]
```

```{r}
xyplot(YEAR ~ Greater_Risk_Data_Value , data = df_youth,
                 main = "Scatterplot Greater_Risk_Data_Value vs year",
                 xlab = "Greater_Risk_Data_Value ",
                 ylab = "YEAR")
```
